Automatic Differentiation Accelerates Inference for Partially Observed Stochastic Processes

### 2024-03-06 presentation/presentation.tex

\begin{frame}{Bias-Variance Tradeoff with $\alpha$}
\begin{figure}
    \centering
    \includegraphics[scale=0.5]{imgs/mlegrad.png}
    \caption{Gradient estimates at MLE for trend in \cite{king08} cholera model. When $\alpha=1$, estimate unbiased with high variance. When $\alpha=0$, lower variance but biased.}
    \label{fig:bias-variance}
\end{figure}

@@

\begin{frame}{Guarantee}
    \begin{prop}[Correctness of MOP-$\alpha$]
    \label{prop:mop-correctness}
    When either $\alpha=1$ or $\theta=\phi$, MOP-$\alpha$ targets the posterior and is unbiased for the likelihood under $\theta$. 
    When $\theta$ is evaluated at $\phi$, the likelihood estimate agrees with the bootstrap filter. Its gradient when $\alpha=1$ is the estimate of \citet{poyiadjis11},
    $$\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}(x_{1:N,j}^{A, F,\theta}, y_{1:N}^*),$$
    and when $\alpha=0,$ is the gradient estimator of \cite{blei2018vsmc},
    $$
        \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta),
    $$
    i.e. differentiating through a vanilla particle filter (\cite{scibior21}).
\end{prop}
\end{frame}

@@

    \begin{table}[h!]
        \centering
        \begin{tabular}{||c c c||} 
         \hline
         Method & Best Log-Likelihood & Rank \\ [0.5ex] 
         \hline\hline
         IFAD-0 & $\textbf{-3747.87}$ & 1\\
         IFAD-1 & $\textbf{-3748.53}$ & 2\\
         IF2 (Ours) & -3755.91 & 3\\
         IF2 (\cite{ionides15}) & -3768.63 & 4\\ 
         MOP-0 Alone & -3795.74 & 5\\
         MOP-1 Alone & -3797.38 & 6\\
         \hline
        \end{tabular}
        \caption{\\
        -- IFAD performs the best among all methods, finding the MLE. \\
        -- Our implementation of IF2 outperforms that of \cite{ionides15} but underperforms IFAD.}
        \label{table:mle}
    \end{table}

    \begin{itemize}
        \item Benchmarked IFAD against IF2 on a challenging global search problem. 
        \item \pause Performed 100 searches each. 
    \end{itemize}

>>>>

\begin{frame}{Bias-Variance Tradeoff with $\alpha$}
\begin{figure}
    \centering
    \includegraphics[scale=0.5]{imgs/mlegrad.png}
    \caption{Gradient estimates at MLE for trend in \cite{king08} cholera model. When $\alpha=1$, estimate close to unbiased with high variance. When $\alpha=0$, lower variance but biased.}
    \label{fig:bias-variance}
\end{figure}

@@

\begin{frame}{Bias-Variance Tradeoff with $\alpha$}
\begin{figure}
    \centering
    \includegraphics[scale=0.5]{imgs/095/biasvar.png}
    \caption{Bias-variance tradeoff for gradient estimates at MLE for trend in \cite{king08} cholera model. When $\alpha=1$, estimate close to unbiased with high variance. When $\alpha=0$, lower variance but biased. MSE seems to be minimized at $\alpha=0.97$.}
    \label{fig:bias-variance}
\end{figure}

\begin{frame}{Guarantee}
    \begin{prop}[Correctness of MOP-$\alpha$]
    \label{prop:mop-correctness}
    When either $\alpha=1$ or $\theta=\phi$, MOP-$\alpha$ targets the posterior and is unbiased for the likelihood under $\theta$. 
    When $\theta$ is evaluated at $\phi$, the likelihood estimate agrees with the bootstrap filter. Its gradient when $\alpha=1$ is the estimate of \citet{poyiadjis11},
    $$\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{0:N}}(y_{1:N}^*|x_{1:N,j}^{A, F,\theta}),$$
    and when $\alpha=0,$ is the gradient estimator of \cite{blei2018vsmc},
    $$
        \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta),
    $$
    i.e. differentiating through a vanilla particle filter (\cite{scibior21}).
\end{prop}
\end{frame}

@@

    \begin{table}[h!]
        \centering
        \begin{tabular}{||c c c||} 
         \hline
         Method & Best Log-Likelihood & Rank \\ [0.5ex] 
         \hline\hline
         IFAD-0.97 & $\textbf{-3750.21}$ & 1\\
         IFAD-0 & $-3752.17$ & 2\\
         IFAD-1 & $-3754.63$ & 3\\
         IF2 (Ours) & -3764.10 & 4\\
         IF2 (\cite{ionides15}) & -3768.63 & 5\\ 
         MOP-1 Alone (100 searches) & -3797.38 & 6\\
         \hline

        \end{tabular}
        \caption{\\
        -- IFAD performs the best among all methods, finding the MLE. \\
        -- Our implementation of IF2 outperforms that of \cite{ionides15} but underperforms IFAD.}
        \label{table:mle}
    \end{table}

    \begin{itemize}
        \item Benchmarked IFAD against IF2 on a challenging global search problem. 
        \item \pause Performed 44 searches each. 
    \end{itemize}
